The Poisson approximate likelihood compared to the particle filter

### 2024-07-25 ms.Rnw
use runif_design, kintable

RUN_LEVEL = 1
NP_MIF       = switch(RUN_LEVEL, 1000, 50000)
NMIF         = switch(RUN_LEVEL, 4,    100)
ncores       = switch(RUN_LEVEL, 2,    36)
top_n_fits   = switch(RUN_LEVEL, 1,    12)
sim_rep      = switch(RUN_LEVEL, 1,    100) # Total number of simulations

#population number
pop <- 82372825
@

@@

<<maximization1, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
set.seed(123)
sir_box <- rbind(
  beta1=c(10,15),
  beta2=c(0.2,0.4),
  beta3=c(0.3,0.5),
  phi=c(0.01,0.3),
  beta11=c(0.1,0.2),
  sigma_q=c(0.001,0.1),
  sigma_xi=c(65,70),
  S10 = c(0.01,0.05),
  I10 = c(0.0001, 0.0005),
  R10 = c(0.01, 0.05),
  S20 = c(0.01,0.7),
  I20 = c(0.000001,0.00002),
  R20 = c(0.001,0.005),
  S30 = c(0.01,0.5),
  I30 = c(0.000001, 0.0001),
  R30 = c(0.0001,0.005)
)

## Make it panelPomp
c(apply(sir_box,1,function(x)runif(1,min=x[1],max=x[2])),
  params_fixed) -> starting

bake(file = paste0("output/round_01/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_01.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_01 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
                    beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
                    S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
                    S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
                    S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting
    ) 
  }
  mifs_global_01
})

bake(file = paste0("output/round_01/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_01.rds"),{
  el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
  el_01
})
@

@@

el_01 <- readRDS(paste0("output/round_01/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_01.rds"))
starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)

bake(file = paste0("output/round_02/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_02.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_02 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
                    beta11=0.01*sd_02,phi=0.01*sd_02,
                    sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
                    S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
                    S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
                    S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_02[[i]]
    ) 
  }
  mifs_global_02
})

bake(file = paste0("output/round_02/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_02.rds"),{
  el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
  el_02
})
@

@@

el_02 <- readRDS(paste0("output/round_02/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_02.rds"))
starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)

bake(file = paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_03.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_03 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
                    beta11=0.01*sd_03,phi=0.01*sd_03,
                    sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
                    S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
                    S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
                    S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_03[[i]]
    ) 
  }
  mifs_global_03
})

bake(file = paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_03.rds"),{
  el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
  el_03
})
@

<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
mif_el <- readRDS(paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_03.rds")) 
# Can be changed if more rounds needed
est_mle_vec <- mif_el[which.max(mif_el$logLik),]
@

@@

bake(file=paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"),{
  ovov_sim_100 <- list()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
bake(file=paste0("output/pfilter/pfilter_ovov_sim_100_",RUN_LEVEL,".rds"),{
  pfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

<<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
prop <- 420

@@

bake(file=paste0("output/pfilter/palfilter_ovov_100_",RUN_LEVEL,".Rdata"),{
  y <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)
  for(i in 1:length(ovov_sim_100)){
    realdat <- ovov_sim_100[[i]] |> as.matrix()
    lik_list <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
                                     c(initial_guess[1],initial_guess[2],initial_guess[3],
                                       initial_guess[4],initial_guess[5]), 
                                     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
                                     norm_par = c(0.07,initial_guess[6]), 
                                     t(prop), NP_MIF,1)
    y[i] <- lik_list$log_lik
  }
  y
})

>>>>

RUN_LEVEL = 1
NP_MIF       = switch(RUN_LEVEL, 10, 50000)
NMIF         = switch(RUN_LEVEL, 4,    100)
ncores       = switch(RUN_LEVEL, 2,    36)
top_n_fits   = switch(RUN_LEVEL, 1,    12)
sim_rep      = switch(RUN_LEVEL, 1,    100) # Total number of simulations

#population number
pop <- 82372825

out_dir <- paste0("output_0",RUN_LEVEL, "/")
@

@@

<<maximization1, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
set.seed(123)
sir_box <- list(
  beta1=c(10,15),
  beta2=c(0.2,0.4),
  beta3=c(0.3,0.5),
  phi=c(0.01,0.3),
  beta11=c(0.1,0.2),
  sigma_q=c(0.001,0.1),
  sigma_xi=c(65,70),
  S10 = c(0.01,0.05),
  I10 = c(0.0001, 0.0005),
  R10 = c(0.01, 0.05),
  S20 = c(0.01,0.7),
  I20 = c(0.000001,0.00002),
  R20 = c(0.001,0.005),
  S30 = c(0.01,0.5),
  I30 = c(0.000001, 0.0001),
  R30 = c(0.0001,0.005),
  #fixed params
  gamma = c(1,1),
  delta1=c(1/(5*52),1/(5*52)),
  delta2=c(1/(55*52),1/(55*52)),
  alpha=c(1/(78.86912*52),1/(78.86912*52)),
  mu=c(0,0),
  N=c(82372825,82372825),
  omega=c(1/(1*52),1/(1*52))
)

design_matrix <- runif_design(lower = sapply(sir_box, `[`, 1),
                              upper = sapply(sir_box, `[`, 2),
                              n = ncores)

# c(apply(sir_box,1,function(x)runif(1,min=x[1],max=x[2])),
#   params_fixed) -> starting

bake(file = paste0(out_dir, "round_01/ovov_mif_01.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_01 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
                    beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
                    S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
                    S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
                    S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = design_matrix[i, ]
    ) 
  }
  mifs_global_01
})

bake(file = paste0(out_dir, "round_01/ovov_el_01.rds"),{
  el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
  el_01
})
@

@@

el_01 <- readRDS(paste0(out_dir,"round_01/ovov_el_01.rds"))
starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)

bake(file = paste0(out_dir,"round_02/ovov_mif_02.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_02 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
                    beta11=0.01*sd_02,phi=0.01*sd_02,
                    sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
                    S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
                    S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
                    S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_02[[i]]
    ) 
  }
  mifs_global_02
})

bake(file = paste0(out_dir, "round_02/ovov_el_02.rds"),{
  el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
  el_02
})
@

@@

el_02 <- readRDS(paste0(out_dir, "round_02/ovov_el_02.rds"))
starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)

bake(file = paste0(out_dir, "round_03/ovov_mif_03.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_03 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
                    beta11=0.01*sd_03,phi=0.01*sd_03,
                    sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
                    S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
                    S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
                    S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_03[[i]]
    ) 
  }
  mifs_global_03
})

bake(file = paste0(out_dir,"round_03/ovov_el_03.rds"),{
  el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
  el_03
})
@

<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
mif_el <- readRDS(paste0(out_dir,"round_03/ovov_el_03.rds")) 
# Can be changed if more rounds needed
est_mle_vec <- mif_el[which.max(mif_el$logLik),]
@

@@

bake(file=paste0(out_dir, "pfilter/ovov_sim_100.rds"),{
  ovov_sim_100 <- list()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/ovov_sim_100.rds"))
bake(file=paste0(out_dir, "pfilter/pfilter_ovov_sim_100.rds"),{
  pfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

<<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/ovov_sim_100.rds"))
prop <- 420

@@

bake(file=paste0(out_dir, "pfilter/palfilter_ovov_100.rds"),{
  y <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)
  for(i in 1:length(ovov_sim_100)){
    realdat <- ovov_sim_100[[i]] |> as.matrix()
    lik_list <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
                                     c(initial_guess[1],initial_guess[2],initial_guess[3],
                                       initial_guess[4],initial_guess[5]), 
                                     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
                                     norm_par = c(0.07,initial_guess[6]), 
                                     t(prop), NP_MIF,1)
    y[i] <- lik_list$log_lik
  }
  y
})
